Matching Bullets

### 2015-12-11 imaging-paper.Rnw
little things

Bullet matching is a process in forensic science done to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of the conviction of criminals in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely describe other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence \eh{XXX we probably need a reference here}.
\hh{XXX have you tried the Exoneration Registry at michigan? http://www.law.umich.edu/special/exoneration/Pages/about.aspx}

@@

The scatterplot in Figure~\ref{fig:residual} shows the residuals of such a fit.
\hh{In this instance, the radius is estimated as $\hat{r} = \Sexpr{pars$radius}\mu m = \Sexpr{pars$radius/1000}mm$ and the land covers about \Sexpr{diff(range(dbr111$theta))} degrees.  Both of these estimates are consistent with a 9 mm bullet fired in a Ruger P-85.}
The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For a surface cross cut at $x = 100$ there is a residual circular structure that does not show up for all cross sections.

%\hh{XXX add parameter estimates? $\hat{r} \approx 4500$ - indicating a 9 mm bullet, amount of degrees covered by land?  }

@@

\begin{enumerate}
    \item Fix an $x$ value (Figure~\ref{fig:loess_step1}, with $x$ set to 243.75$\mu m$).
    \item For each $y$ value, smooth out any deviations occurring near the minima by twice applying a rolling average with a pre-set \textbf{smoothing factor}. (Figure~\ref{fig:loess_step3}, smoothing factor of 35, corresponding to 55$\mu m$).
%    \item For each smoothed $y$ value, compute another rolling average using the same smoothing factor as above. (Figure~\ref{fig:loess_step3}).
    \item Determine the location of the peak of the left groove by finding the first double-smoothed value $y_i$ that is the maximum within its smoothing window (e.g.\ such that $y_i > y_{i - 1}$ and $y_i > y_{i + 1}$, $i = 1, ..., 17$). We call the location of this peak $p_{left}$ (see Figure~\ref{fig:loess_step47}). 
    %\hh{XXX By peak, do you mean the location of the peak or its height or both?} XXX never mind, found the answer below
    \item Similarly, determine the location of the valley of the left groove by finding the first double-smoothed $y_j$ that is the minimum within its smoothing window. Call the location of this valley $v_{left}$.
    \item Reverse the order of the $y$ values and repeat the previous two steps to find the peak and valley of the right groove, $(p_{right}, v_{right})$.
    \item Trim the surface measurements to values within the two grooves (i.e.\ remove all records with $y_i < v_{left}$ and $y_i > v_{right}$). (Figure \ref{fig:loess_step47}).
\end{enumerate}

@@

\subsection{Varying smoothing factor}
As mentioned, the algorithm for peak/valley detection depends on the selection of a smoothing window, called the smoothing factor. In particular, for detecting and removing the grooves prior to fitting a Loess regression, we selected a smoothing factor of 35. We can investigate the performance of groove detection across different smoothing factors. Figure \ref{fig:varysmooth}...

\begin{figure}[hbtp]
\centering
\begin{subfigure}[t]{\linewidth}
\caption{Groove detection for bullet 1-5 at a smoothing factor of 35.\label{fig:smoothfac1}}{%
<<smoothfac1, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
get_grooves(br111_bullet, smoothfactor = 35, smoothplot = TRUE)$plot
@
    }
\end{subfigure}    
\begin{subfigure}[t]{\linewidth}
    \caption{Groove detection for bullet 1-5 at a smoothing factor of 5.\label{fig:smoothfac2}}{%
<<smoothfac2, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
get_grooves(br111_bullet, smoothfactor = 5, smoothplot = TRUE)$plot
@
    }
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
    \caption{Groove detection for bullet 1-5 at a smoothing factor of 65.\label{fig:smoothfac3}}{%
<<smoothfac3, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
get_grooves(br111_bullet, smoothfactor = 65, smoothplot = TRUE)$plot
@
    }
\end{subfigure}
\caption{\label{fig:varysmooth} Groove detection at smoothing factors of 35, 5, and 65 respectively. Note that a smoothing factor of 5 yields enough variation prior to the groove that the right groove is not detected properly, while a smoothing factor of 65 over-smoothes and causes the peak on the right to disappear, so that the peak and groove are both not detected properly.}
\end{figure}

@@

Figure~\ref{fig:match} gives an overview of the automated matching routine: 
We first identify a stable region for each bullet and pick the lowest crosscut level in this region \hh{(assumption here: lower crosscut levels show better expression of individual marks XXX need to still check)}

>>>>

Bullet matching is a process in forensic science done to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of the conviction of criminals in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely describe other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence \cite{giannelli:2011}.

@@

The scatterplot in Figure~\ref{fig:residual} shows the residuals of such a fit.
In this instance, the radius is estimated as $\hat{r} = \Sexpr{pars$radius}\mu m = \Sexpr{pars$radius/1000}mm$ and the land covers about \Sexpr{diff(range(dbr111$theta))} degrees.  Both of these estimates are consistent with a 9 mm bullet fired in a Ruger P-85.
The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For a surface cross cut at $x = 100$ there is a residual circular structure that does not show up for all cross sections.

@@

\begin{enumerate}
    \item Fix an $x$ value (Figure~\ref{fig:loess_step1}, with $x$ set to 243.75$\mu m$).
    \item For each $y$ value, smooth out any deviations occurring near the minima by twice applying a rolling average with a pre-set \textbf{smoothing factor}. (Figure~\ref{fig:loess_step3}, smoothing factor of 35, corresponding to 55$\mu m$).
%    \item For each smoothed $y$ value, compute another rolling average using the same smoothing factor as above. (Figure~\ref{fig:loess_step3}).
    \item Determine the location of the peak of the left groove by finding the first double-smoothed value $y_i$ that is the maximum within its smoothing window (e.g.\ such that $y_i > y_{i - 1}$ and $y_i > y_{i + 1}$, $i = 1, ..., 17$). We call the location of this peak $p_{left}$ (see Figure~\ref{fig:loess_step47}). 
    \item Similarly, determine the location of the valley of the left groove by finding the first double-smoothed $y_j$ that is the minimum within its smoothing window. Call the location of this valley $v_{left}$.
    \item Reverse the order of the $y$ values and repeat the previous two steps to find the peak and valley of the right groove, $(p_{right}, v_{right})$.
    \item Trim the surface measurements to values within the two grooves (i.e.\ remove all records with $y_i < v_{left}$ and $y_i > v_{right}$). (Figure \ref{fig:loess_step47}).
\end{enumerate}

@@

\subsection{Varying smoothing factor}
As mentioned, the algorithm for peak/valley detection depends on the selection of a smoothing window, called the smoothing factor. In particular, for detecting and removing the grooves prior to fitting a Loess regression, and for detecting the peaks/valleys of the loess residuals, we selected a smoothing factor of 35. We can investigate the performance of peak/valley detection across different smoothing factors. Figure \ref{fig:varysmooth} displays the resulting peaks and valleys across smoothing factors of 35, 5, and 65 respectively. It is immediately clear that a small smoothing factor like 5 is a poor choice. This results in a significant amount of noise in the data such that even just a point or two can skew the rolling average enough for a peak or valley to be detected. Given that the striation patterns are typically much larger, we're in effect muddying the waters by performing such minimal smoothing.

A larger smoothing factor on the other hand, like 65, seems to be a more plausible option. Most of the peaks/valleys present which are detected by a smoothing factor of 35 are also detected at 65. However, some notable issues arise. Notice that the peak on the left hand side of the image is smoothed out, and thus not detected. On the right hand side, the peak is just barely still present and hence detected. But in many cases, the large windows are smoothing out some of the structure that we wish to see. Furthermore, it can be seen that the peaks/valleys are often shifted relative to their position in the original loess residuals, or in the smoothed data with smaller smoothing factors.

\begin{figure}[hbtp]
\centering
\begin{subfigure}[t]{\linewidth}
\caption{Peak and valley detection for bullet 1-5 at a smoothing factor of 35.\label{fig:smoothfac1}}{%
<<smoothfac1, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
br111.groove <- get_grooves(br111_bullet)
br111.loess <- fit_loess(br111_bullet, br111.groove)
get_peaks(br111.loess$data, smoothfactor = 35)$plot
@
    }
\end{subfigure}    
\begin{subfigure}[t]{\linewidth}
    \caption{Peak and valley detection for bullet 1-5 at a smoothing factor of 5.\label{fig:smoothfac2}}{%
<<smoothfac2, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
get_peaks(br111.loess$data, smoothfactor = 5)$plot
@
    }
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
    \caption{Peak and valley detection for bullet 1-5 at a smoothing factor of 65.\label{fig:smoothfac3}}{%
<<smoothfac3, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
get_peaks(br111.loess$data, smoothfactor = 65)$plot
@
    }
\end{subfigure}
\caption{\label{fig:varysmooth} Peak/valley detection at smoothing factors of 35, 5, and 65 respectively. Note that a smoothing factor of 5 yields enough noise that many very minimal overlapping peaks and valleys are detected, while a smoothing factor of 65 over-smoothes and causes the peaks/valleys to either disappear or be horizontally shifted from their position in the original loess residuals.}
\end{figure}

@@

Figure~\ref{fig:match} gives an overview of the automated matching routine: 
We first identify a stable region for each bullet and pick the lowest crosscut level in this region, because typically, individual characteristics are best express at the lower end of the bullet. 
%\hh{(assumption here: lower crosscut levels show better expression of individual marks XXX checked XXX crosscuts-and-sd.R)}
